Molecular Gas Kinematics of the Strongly-lensed Quasar Host Galaxy RXS J1131$-$1231

### 2016-11-17 WRITEUP/response.md

We have fixed the number of decimals in spectral indices in Sec 3.3.

10) We have replotted the figure with different contour intervals with thinner lines and added a sentence in the caption to improve clarify of Fig 4.

11) We have corrected "blueshifted" to "redshifted", udpated the velocity offset from 780 km/s to 715 km/s using z = 0.65406.

12) 
a) We have removed the statement regarding "AGN dynamically offset".

Our CO measurement confirms that an offset between BLR and the systemic redshift, which is not known previously. We have updated this section, adding additional possible interpretation for the offset (e.g. gas inflows, geometric effects) in the discussion.

We agree that it is unlikely that all objects with offset AGN lines are black hole recoils. Here, we are not asserting that the offset arises due to a recoiling black hole, we only discuss the possibility of such interpretation given the data (showing evidence of large offset between BLR and the systemic redshift as traced by our CO data), the high spin parameter, and theoretical expectation from hierarchical model of galaxy evolution, among other possible scenarios, as listed by the referee.

We would greatly appreciate the referee if they can point us to a reference where it shows evidence of why the recoiling black hole scneario is highly unlikely and will gladly make the appropriate changes.

3) Assuming that effects (e.g., inflow, geometric effect, micro-lensing) are accountable for part of the velocity offset, then the velocity offset of ~715 km/s would then be at most an upper limit on the line-of-sight velocity. Then  the true recoil velocity is thus most likely to be << 1000 km/s. Ignoring dynamical friction against stellar background and assuming t_recoil ~ 100 Myr, then the crude BH displacement estimate given in the referee report (~few 10kpc) would be possible (as pointed out to be theoretically possible 
by Bonning+07, Loeb07). However, Loeb07 also pointed out that at low ejection speeds (few hundred km/s), the deceleration of the BH (damped) by the external gravitational potential of the galaxy and by dynamical friction (Chandrasekhar 1943; Apj 97, 255) on the background stars, dark matter, and gas would limit the displacement of the slow-moving BH and tend to bring it back to the center of its host galaxy within a few dynamical times (see also Madau & Quataert 04 for analytic results and Boylan-Kolchin+04 for numerical results). Therefore the displacement given in the referee report is unlikely to hold (see also Guedes+11).

Finally, again, we acknowledge that offsets between BLR and NLR lines are quite common in Seyferts and quasars, but typically not as big of an offset. In this paper, we only discuss possible scenarios and do not assert that any one of mentioned scenario is necessary in RXJ1131.

### 2016-11-17 WRITEUP/ms.tex
collected arguments regarding the recoil BH section

\subsection{Systemic Redshift and Velocity Offset}
%Outflow:
% Not AGN outflows because require different velocity offsets in Balmer and forbidden lines (Zamanov02a)
% line origins:
% \mgii & Hbeta: outer part of the BLR (Goad12a) compared to high-ionization lines (\eg CIV)
% Hbeta can also come from NLR

\citet[]{Sluse07a} report two sets of AGN lines observed in RXJ1131.
The first set of lines is at $z$\ssim0.654, including the narrow component of the Balmer lines, the \oiii lines, and the \mgii absorption line; the second set is at $z$\ssim0.658, including the broad component of the Balmer lines and
the \mgii emission line.
Using the CO line center redshift as the systemic redshift,
we find that the redshift of the first set is fully consistent with the systemic redshift. This
supports previous claims that [O\,{\scriptsize III}] lines, tracing the narrow line region (NLR),
are good proxies
to the true systemic redshift \citep[e.g.,][]{Vrtilek85a, Nelson00a}.
On the other hand, the second set of lines is redshifted by $\sim$715\,\kms.

% REFEREE: offset in BLR and NLR in RXJ1131 has been known before, not new for this paper
%One explanation for the offset may arise from the fact that BLRs are gravitationally bound to
%the central AGN, whereas this condition is not necessary for NLRs.
%From an SDSS sample of $>$3800 quasars, \citet{Richards02a} find a
%median of 97\kms and a dispersion of 269\,\kms in the distribution of
%the offset between [Mg\,{\scriptsize II}] and [O\,{\scriptsize III}] lines, suggesting
%that large offsets are uncommon. Nevertheless,
%offsets as large as $\sim$\,500\,\kms have been observed in some cases.
%%% In an SDSS sample of ongoing mergers at $z$\,$<$\,0.21, AGN lines are offset by $\lesssim$410\,km/s
%%\citep[\eg][]{Richards02a, Comerford14a}. % between AGN lines (forbidden, Balmer) and stellar absorption

%Alternatively, large velocity offsets between emission lines of the
%BLR and the NLR may arise in a recoiling black hole (BH),
%% recoiling SMBHs as well as subparsec-scale binary SMBHs produce velocity offsets in the broad AGN emission lines and not the narrow AGN emission lines (e.g., Gaskell 1983; Bonning et al. 2007; Eracleous et al. 2012; Ju et al. 2013; Shen et al. 2013).
%where its BLR is moving at high velocity relative to the bulk of its host galaxy
%\citep{Madau04a, Bonning07a, Loeb07a}.
%A recoiling black hole is expected
%when a pair of uneven mass BH coalesce, during which their orbital energy is being released
%as gravitational wave and a
%non-zero net angular momentum is being carried away.
%Depending on their initial conditions, numerical relativity simulations have shown that
%the recoil velocity can reach up to $\sim$4000\,\kms for spinning BHs \citep[\eg][]{Campanelli07a}.
%Several sources have been proposed as recoiling BH candidates, such as CID-42 at $z$\,=\,0.359 \citep{Civano10a} and
%SDSS 0956+5127 at $z$\,$\sim$\,0.714 \citep{Steinhardt12a}.
%Since this model requires a merged BH, it would imply that RXJ1131 might
%have been in a previous merger, which would also explain
%the highly spinning BH in RX1131 \citep[$a$\ssim0.9;][]{Reis14a}.

>>>>

\subsection{Systemic Redshift and Velocity Offset}
%Outflow:
% require different velocity offsets in Balmer and forbidden lines (Zamanov02a)
% line origins:
% \mgii & Hbeta: outer part of the BLR (Goad12a) compared to high-ionization lines (\eg CIV)
% Hbeta can also come from NLR

\citet[]{Sluse07a} report two sets of AGN lines observed in RXJ1131.
The first set of lines is at $z$\ssim0.654, including the narrow component of the Balmer lines, the \oiii lines, and the \mgii absorption line; the second set is at $z$\ssim0.658, including the broad component of the Balmer lines and the \mgii emission line.
Using the CO line center redshift as the systemic redshift,
we find that the redshift of the first set is fully consistent with the systemic redshift. This
supports previous claims that [O\,{\scriptsize III}] lines, tracing the narrow line region (NLR),
are good proxies
to the true systemic redshift \citep[e.g.,][]{Vrtilek85a, Nelson00a}.
On the other hand, the second set of lines is redshifted by $\sim$715\,\kms.
Offset between BLR and NLR lines have been reported in literature, e.g., 
\citet{Richards02a} find a
median of 97\kms and a dispersion of 269\,\kms in the distribution of
the offset between [Mg\,{\scriptsize II}] and [O\,{\scriptsize III}] lines in an SDSS sample of $>$3800 quasars, 
and \citet{Bonning07a} find a mean offset of 100\,\kms with a dispersion of 212\,\kms %FWHM = 500
between the broad component of
H$\beta$ and \oiii lines in a sample of $\sim$2600 quasars at 0.1$<$$z$$<$0.81,
illustrating that large offsets ($\Delta v$$\gtrsim$\,500\,\kms) are uncommon but have been observed in some cases.
%% In an SDSS sample of ongoing mergers at $z$\,$<$\,0.21, AGN lines are offset by $\lesssim$410\,km/s
%\citep[\eg][]{Richards02a, Comerford14a}. % between AGN lines (forbidden, Balmer) and stellar absorption

Common interpretations used to explain the offset between optical and systemic redshift include
e.g., outflow/inflow of gas in the BLR, 
where the dominant component of the BLR linewidth
is the rotation of the accretion disk, superimposed by an large inflow component, 
or arising from the inclination angle of the (inner) accretion disk, where 
clumps in the disk that is more edge-on block the far side, and thus only see the redshifted (for RXJ1131) 
inflowing component on the BLR gas.
{\bf DR: is it because of the large linewidth (one component of \mgii is $>$20000\kms, and broad comp. of H$\beta$ is $\sim$2000$-$4000\kms) that the referee thinks it's more likely to be inflow and therefore doesn't like the recoil black hole?}
Additionally, microlensing in RXJ1131 \citep[\eg][]{Sluse07a, Sluse12a} may preferentially magnified the (red wing of the compact BLR 
emission), and thus biasing the BLR line centroids.% , i.e., the redshift offset may not be a real one.
% Sluse07: BLR with wide FWHM (> 2000\kms in \mgii and H$\beta$), one component of \mgii > 20000 \kms --> partial attentuation of the blue sides of the line could shift the apparent BLR line centroid to the red side.

%  --------------- note --------------
% recoil...
% Given that there's an accretion disk (Dai+10, Pooley+07).., which could have been reform after the merging (in ~100 year; Bonning+07), then it would mean that the binary merged > blah years (presumably from previous encounter).
% [ignore this] could also be currently a binary orbiting at high speed? (do we see multiple components in the BLR lines? Do we know that there's only a single BH from Xray? probably only 1 AGN, because otherwise we would see more complicated lensed QSO emission)
%
% Sluse07:  "This indicates that the flux in the NEL is partly resolved along the slit direction. On the other hand, we did not find a significant offset (i.e. offset <1 pixel) of the centroids of the NEL images w.r.t. to the centroid of the continuum emission. Using a SIE+? to model the lensing galaxy (Paper I), we calculate that the absence of such an offset implies an upper limit of roughly ?30h?1 pc on the off-centering of the [O III] emission w.r.t. the continuum emission."
% --> so the NLR is only slightly spatially offset from the continuum, and therefore must be inflow and microlensing? 
% --> but couldn't it still be recoil even if it's not spatially offset (perhaps unlikely?) e.g. Allen+15, Section 4.1 "the [O III] peaks do not coincide spatially with the X-ray sources, indicating that they may not be directly due to the two AGN, and suggest a bi-conical outflow as a likely alternative." << sounds like if the NLR is not co-spatial with the AGN, then could be outflow. whereas if referee thinks no spatial offset (Sluse+07) and therefore probably inflow rather than recoil AGN doesn't seem to hold. 
% ----> hmm, but I guess outflow could be spatially offset but not necessary for inflow.
% --------------------------------------

In hierarchical model of galaxy formation, supermassive black holes are formed by
mergers, of which the majority took place at high redshifts. 
% The most commonly used technique to find offset AGNs is to 
% look for shifts between BLRs associated with the QSO and the NLR associated 
% with the starlight from the galaxy. 
We also discuss the possibility of a recoil black hole from previous episode of merger at earlier epoch \citep{Madau04a} leading to the offset... 
Alternatively, the offset between the BLR and the systemic redshift 
may be partially due to a kinematically offset AGN \citep{Hoffman & Loeb (2006)}, where
BLRs are gravitationally bound to
the central AGN (with a BLR dynamical timescale $\sim$10\E{2} years \citep{Bonning07a}, \ie can be quickly regenerated), 
whereas this condition is not necessary for NLRs.
% But Madau04a: a ~1e8\Msun BH in a $\sigma$\eq200\kms galaxy, the decay timescale due to dynamical friction is $<$ 1 Myr $<<$ Salpeter timescale, and so even for a kick velocity of v_CM\eq300\kms, the displacement is quite small $\lesssim$$R_{\rm BH}$\approx$10 pc.

Alternatively, large velocity offsets between emission lines of the
BLR and the NLR may arise in a recoiling black hole (BH),
where its BLR is moving at high velocity relative to the bulk of its host galaxy
\citep{Madau04a, Bonning07a, Loeb07a}.
A recoiling black hole is expected % once they overcome the ``final pc problem'' by e.g. stellar dynamical hardening/hardening by 
% dynamical friction against background stars, or viscous gas dynamical drag.
when a pair of uneven mass BH coalesce, during which their orbital energy is being released
as gravitational wave and a non-zero net angular momentum is being carried away.
Depending on their initial conditions (e.g., black hole mass ratio, spin-orbit orientation, spin magnitude), numerical relativity simulations have shown that the recoil velocity can reach up to  $v_{\rm kick}$\ssim4000\,\kms for spinning BHs \citep[\eg][]{Campanelli07a}, 
with a typical offset recoil velocity $v_{\rm kick}$ in the range of 100$-$500\,\kms 
% based on simulations, and simulation constraints from the scatter in the local $M_{\rm BH}$$-$\mbulge relation
\citep[\eg][]{Favata04a,Libeskind06a,Campanelli07a}.
The black hole can therefore either leave the host galaxy (if vkick $>$ $v_{\rm esc}$) or 
oscillate around the center of the galaxy, its orbit decaying because of dynamical friction 
\citep[see \eg \citealt{Volonteri10a} for a discussion]{Bonning07a, Colpi14a}.

% escape velocity
$M_{\rm BH}$/\mstar\ssim1/375, $v_{\rm circ}$\ssim$v_{\rm rot}, then the upper limit of $v_{\rm kick}$\ssim2$\times$$v_{\rm circ}$, 
assuming bulge escape velocity $v_{\rm esc}$\eq2.83$\v_{\rm circ}$ = 858\kms
(in the case of a density profile of the spherical \citet{Hernquist90a} model for the stellar bulge component, 
spherically symmetric isotropic model)%  \citep{BoylanKolchin04a}, 
$v_esc$\eq2sqrt(2) *  $v_{\rm circ}$($r$\eq$a$), where r is the stellar bulge scale length, at most a few kpc, 
therefore finding $v_esc$\eq BLAH\kms.
$t_dyn$ would be 
% Assuming know the rotation curve of RXJ1131 is similar to that MW, then at inner radius, we would expect $v_{\rm rot} (r=a)\gtrsim$ the extended 
% gas disk rotation, given that we don't know precisely the stellar bulge scale length, if we assume 
% $v_{\rm rot} (r=a)$\ssim$v_{\rm rot}$ of the gas disk, then v_esc ~ 860\kms. Therefore BH unlikely to escape anyway.

In the case of gas-rich merger remnants, where large amounts of gas have been
funneled to the central regions find that the MBHs remain within R$<$1\,kpc from the center
of the host even for high recoil velocities ($v_{\rm kick}$\eq1200\,\kms) due to the compactness of the
remnant galaxy?s nuclear disk" \citep{Mayer07a,Guedes08a}.
Several outcomes of the recoiling blackholes include,
a) In spherical bulges the black hole will oscillate right through the center, and return to the center after a few dynamical times, or 
deplete or "heat up" the stars in the nucleus, reducing the efficiency of dynamical and extending the return time.
% in gas-rich mergers, gas drag, duty cycle of the wandering AGN..
% depending on the nuclear disk density profile, the MBH is unlikely to reach large distances
%  the timescale for the recoiling hole to return to the center of its host galaxy depends very sensitively on the magnitude of the kick it receives, ranging from $\sim$10\E{6} yr for v_CM\eq200 \kms to more than 109 yr for v_CM \eq 400 \kms \citep{Madau04a} 
% Loeb07, Madau04a, Hoffman07: At low ejection speeds, the deceleration of the BH by the external gravitational potential of the galaxy and by dy- namical friction on the background stars and gas would limit the displacement of the slow-moving BH and tend to bring it back to the center of its host galaxy within a few dynamical times
b) If the black hole is able to leave the bulge / ejected and enters a triaxial dark matter halo, it may not return at all within a Hubble time \citep{Guedes09a} and become intergalactic/naked black holes (but rare unless in recoil velocity $>>$ few thousands \kms, i.e., sufficient to escape the host galactic potential) \citep[\eg][]{Loeb07a}.
% Volonteri+10: ejected BHs are rare at $z$$<$5, especially in wet-mergers
%c) High resolution simulations of gas mergers (e.g. Mayer et al. 2007) show that up to 60\% of the initial gas is funneled to the center of the remnant. In that particular simulation, the mass of the nuclear disk is Md = 3e9 Msun, and recoliing black holes with velocities of 
%$~$500 km/s are not able to escape the central region (scale length L=75 pc).

Several sources have been proposed as recoiling BH candidates, such as CID-42 at $z$\,=\,0.359 \citep{Civano10a},
SDSS 0956+5127 at $z$\,$\sim$\,0.714 \citep{Steinhardt12a}, and
SDSSJ0927+2943 at $z$\,$\sim$\,0.713, where a set of redshifted narrow lines and a second set of blueshifted broad AGN lines
have been observed \citep{Komossa08a}.
However, this scenario for SDSSJ0927+2943 was recently refuted by \citet{Decarli14a}, who find that the
BLR lines in SDSSJ0927+2943 are indeed consistent with the CO systemic redshift.
This is in contrast with RXJ1131, where our CO observations confirm the redshifted BLR.
Since this model requires a merged BH, it would imply that RXJ1131 might
have been in a previous merger, which might explain the highly spinning BH in RX1131 \citep[$a$\ssim0.9;][]{Reis14a}. 
% will spin up if major merger but would spin down in minor merger \citep{Gammie04a}, in general would spin down after merging.
% Dave Chernoff: can spin up after merging
The available data do not provide direct evidence favoring any of the above scenarios, 
we therefore cannot rule out any of the above plausible scenarios; $\simless$ 1-kpc
resolution observations probing the morphology, structures and kinematics of the (ionized) gas and stellar (e.g., via H$\alpha$ kinematics measurements) would be useful to distinguish between these scenarios.  
% I think sluse+07 constrained the size of BLR based on model and not really resolving down to those spatial resolution? e.g. not really see outflows
